\bf Cognitive control and semantic thought variability\\across sleep and wakefulness

### 2024-12-15 manuscript/manuscript.tex
URLs as citations

\begin{figure*}[t]
    \centering
    \includegraphics{figure1.pdf}
    % \caption{\textbf{Semantic incoherence as a measure of thought variability.} Thought variability is measured by generating a single semantic representation of each thought, finding the semantic distance between each consecutive thought pair, and averaging those distances together. Shaded boxes represent relative numerical values and are for illustrative purposes only. Example text is a haiku written by Tracy Davidson~\cite{tracy2021}.}
    \caption{\textbf{Semantic incoherence as a measure of thought variability.} Thought variability is measured by generating a single semantic representation of each thought, finding the semantic distance between each consecutive thought pair, and averaging those distances together. Shaded boxes represent relative numerical values and are for illustrative purposes only. Example text is a haiku written by Tracy Davidson\protect\footnotemark.}
    \label{fig:methods}
\end{figure*}

@@

\subsection*{Waking-thoughts dataset}
To capture waking thoughts under variable levels of cognitive control in daily life, we used experience samples from an ecological momentary assessment study~\cite{mills2021}. Participants reported their thoughts up to 6 times per day for one week in response to text message prompts. Participants were told---upon receiving a phone alert---to ``take a mental snapshot of what they were thinking about just before they saw the probe'' and then answer some questions regarding that mental snapshot. They were asked: (1) A free-text description of their ongoing thoughts: ``Please provide a brief description of your most recent stream of thought based on the mental snapshot.'' (2) A report on task-unrelatedness, or whether those thoughts were on- or off-task: ``Are you thinking about something other than what you are currently doing? Answer with either a \textit{Y} or an \textit{N}.'' (3) A report on a 6-point Likert scale about freedom of movement in thought, or how much those thoughts were freely moving: ``Are your thoughts wandering around freely? Answer on a scale from 1 (\textit{not at all}) to 6 (\textit{very much so}).'' The second and third measures have been used to assess mind wandering in previous studies and are tied to constructs of constraints and executive control~\cite{mills2018a}. See Study 1 of Mills et al.~\cite{mills2021} for additional details regarding this dataset, including sample information and additional measures that were collected. After text preprocessing (see below), the final sample included $46$ participants. For the comparison of on- and off-task thoughts, the sample was further reduced to participants who reported both $\ge 1$ on-task thought and $\ge 1$ off-task thought ($\mathit{N}=44$).
\footnotetext{\url{https://classicalpoets.org/2021/09/05/the-21-best-haiku-of-2021}}
% \footnotetext{\url{https://classicalpoets.org/2021/07/28/haiku-competition-2021/#comment-479800}}

\subsection*{Dream dataset}
To capture dreams (i.e., sleeping thoughts) under variable levels of cognitive control, we used dream reports from a public online dream journal\footnote{\url{https://www.dreamviews.com/blogs}}~\cite{schredl2022}. All dream-journal entries on this site include the option to tag the dream as lucid and/or nonlucid. We started with all dream reports where the author specified that the dream was either only lucid or only nonlucid, and then reduced the dataset further to only include authors who reported at least one or more of both dream types. After subsequent text preprocessing (see below), the final sample was reduced to authors who reported both $\ge 1$ lucid dream and $\ge 1$ nonlucid dream ($\mathit{N}=1199$).

>>>>

\begin{figure*}[t]
    \centering
    \includegraphics{figure1.pdf}
    \caption{\textbf{Semantic incoherence as a measure of thought variability.} Thought variability is measured by generating a single semantic representation of each thought, finding the semantic distance between each consecutive thought pair, and averaging those distances together. Shaded boxes represent relative numerical values and are for illustrative purposes only. Example text is a haiku written by Tracy Davidson~\cite{tracy2021}.}
    \label{fig:methods}
\end{figure*}

@@

\subsection*{Waking-thoughts dataset}
To capture waking thoughts under variable levels of cognitive control in daily life, we used experience samples from an ecological momentary assessment study~\cite{mills2021}. Participants reported their thoughts up to 6 times per day for one week in response to text message prompts. Participants were told---upon receiving a phone alert---to ``take a mental snapshot of what they were thinking about just before they saw the probe'' and then answer some questions regarding that mental snapshot. They were asked: (1) A free-text description of their ongoing thoughts: ``Please provide a brief description of your most recent stream of thought based on the mental snapshot.'' (2) A report on task-unrelatedness, or whether those thoughts were on- or off-task: ``Are you thinking about something other than what you are currently doing? Answer with either a \textit{Y} or an \textit{N}.'' (3) A report on a 6-point Likert scale about freedom of movement in thought, or how much those thoughts were freely moving: ``Are your thoughts wandering around freely? Answer on a scale from 1 (\textit{not at all}) to 6 (\textit{very much so}).'' The second and third measures have been used to assess mind wandering in previous studies and are tied to constructs of constraints and executive control~\cite{mills2018a}. See Study 1 of Mills et al.~\cite{mills2021} for additional details regarding this dataset, including sample information and additional measures that were collected. After text preprocessing (see below), the final sample included $46$ participants. For the comparison of on- and off-task thoughts, the sample was further reduced to participants who reported both $\ge 1$ on-task thought and $\ge 1$ off-task thought ($\mathit{N}=44$).

\subsection*{Dream dataset}
To capture dreams (i.e., sleeping thoughts) under variable levels of cognitive control, we used dream reports from a public online dream journal~\cite{dreamviews,schredl2022}. All dream-journal entries on this site include the option to tag the dream as lucid and/or nonlucid. We started with all dream reports where the author specified that the dream was either only lucid or only nonlucid, and then reduced the dataset further to only include authors who reported at least one or more of both dream types. After subsequent text preprocessing (see below), the final sample was reduced to authors who reported both $\ge 1$ lucid dream and $\ge 1$ nonlucid dream ($\mathit{N}=1199$).
